Rotation and Lithium Confirmation of a 500 Parsec Halo for the Open Cluster NGC\,2516\footnote{This is paper 3 of the Cluster Difference Imaging Photometric Survey.}

### 2021-04-27 cluster_paper/ms.tex
add literature comparison
Co-authored-by: Joel Hartman <jhartman@astro.princeton.edu>

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.7\textwidth]{f15.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Masses and absolute magnitudes for the innermost and
  outermost stars in \cn.} The \nkinematic\ candidate \cn\ members
  were split into three groups according to their distance from the
  cluster center.  The inner and outermost groups, each containing
  1,099 stars, are counted in 1\,mag bins.  The mean star in the
  innermost sample has a 10\% lower mass than the mean star in the
  outermost sample, consistent with expectations for tidal tails.
  \label{fig:massfunction}
	}
\end{figure*}

@@

From their rotation period analysis of \cn,
\citet{fritzewski_rotation_2020} demonstrated that
between 100 and 200\,Myr, young clusters (\cn, Blanco\,1, Psc-Eri,
M\,35, M\,50, and the Pleiades) all show similar rotation period
distributions.  The FGK stars show a ``slow sequence'', a population
of ``rapid rotators'', and in some cases stars in the ``void'' between
the two populations ({\it e.g.,}, Psc-Eri and \cn\ show more stars in
the void than the Pleiades).  The late M dwarfs with detected rotation
periods tend to rotate rapidly ($<3$\,days), though there may also be
late M dwarfs with slow ($>20$\,days) rotation periods \citep{stauffer_rotation_2016}.  Although
these late M dwarfs are beyond the magnitude limit of our analysis,
they have been explored in depth for \cn\ by both
\citet{Irwin_NGC2516_2007} and \citet{fritzewski_rotation_2020}.
A detailed comparison of our results with theirs is beyond the
scope of this work, but we emphasize that the main contribution of our
study beyond these previous analyses is our sensitivity
across both the core and halo of \cn, and a
corresponding expansion in the number of stars that can be analyzed.
In the following section, we therefore focus on the origins of the
features in the rotation-color diagram, and whether our expanded
sample can provide any new insights.

@@

The other features of the rotation-color diagram are indications that
magnetic braking is not always the only important effect.  The
question of why the rapid rotators exist, for instance, requires
additional physics.  Models that vary core-envelope decoupling
timescales and pre-MS disk lifetimes can reproduce each type of
population
\citep{Irwin_NGC2516_2007,gallet_improved_2013,gallet_improved_2015};
the question then shifts to why these timescales should differ for
stars that are otherwise identical.  Another possibility, perhaps
favored, is that proposed by \citet{matt_mass-dependence_2015}: the
saturation of the magnetic dynamo yields $\dot{P}\sim P$ when $P >
P_{\rm sat}$, $\dot{P} \sim P^{-1}$ for $P < P_{\rm sat}$, and a
maximal spin-down rate in the intermediate regime.  This could
naturally produce the period bimodality observed at ages of
$\approx$100\,Myr (though it fails to explain the widening of the
distribution observed at later times
\citealt{godoyrivera_stellar_2021}).  Other significant processes
could be external to the star entirely: \citet{qureshi_signature_2018}
for instance have reported that mergers between giant planets and the
star could explain a non-neglible fraction of the rapid rotators.

@@

On the other hand, our ability to reliably identify members of these
halos is rather surprising.  These stars have remained hidden in the
background of the Galaxy throughout centuries of modern astronomy.
Their discovery likely represents a major step toward understanding
the dispersal of open clusters.
Combined with wide-field photometric surveys, the stars themselves
enable advances in studies of stellar
angular momentum evolution.
Combined with wide-field spectroscopic
surveys \citep[{\it e.g.},][]{kollmeier_2017}, the halos could also
improve our
understanding of the chemical evolution of low-mass stars.
Finally, the halo stars provide an expanded sample of
age-dated stars to search for planets \citep[{\it
e.g.},][]{newton_2021}, which may help in benchmarking our understanding of
planetary evolution.

@@

The authors thank, in reverse-alphabetical order, G.~Zhou,
B.~Tofflemire, A.~McWilliam, E.~Newton, L.~Hillenbrand, A.~Kraus, and
K.~Hawkins for the discussions on young stars, rotation, and lithium
that encouraged this analysis.
%
L.G.B. and J.H. acknowledge support by the TESS GI Program, programs
G011103 and G022117, through NASA grants 80NSSC19K0386 and
80NSSC19K1728.
%
L.G.B. was also supported by a Charlotte Elizabeth Procter Fellowship
from Princeton University.
%
This study was based in part on observations at Cerro Tololo
Inter-American Observatory at NSF's NOIRLab (NOIRLab Prop{.} ID
2020A-0146; 2020B-0029 PI: L{.}~Bouma), which is managed by the
Association of Universities for Research in Astronomy (AURA) under a
cooperative agreement with the National Science Foundation.
%
% ACKNOWLEDGE PFS / CAMPANAS.
%
This paper also includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
We thank the TESS Architects (G.~Ricker, R.~Vanderspek, D.~Latham,
S.~Seager, J.~Jenkins) and the many TESS team members for their
efforts to make the mission a continued success.
%

>>>>

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f15.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Masses and absolute magnitudes for the innermost and
  outermost stars in \cn.} The \nkinematic\ candidate \cn\ members
  were split into three groups according to their distance from the
  cluster center.  The inner and outermost groups, each containing
  1,099 stars, are counted in 1\,mag bins.  The mean star in the
  innermost sample has a 10\% lower mass than the mean star in the
  outermost sample, consistent with expectations for tidal tails.
  \label{fig:massfunction}
	}
\end{figure}

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.95\textwidth]{f16.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Comparison against previous analyses of stellar
  rotation in \cn.} 
  The bottom panel is a concatenation of Set $\mathcal{B}$ with the
  results of \citet{Irwin_NGC2516_2007} (I07) and
  \citet{fritzewski_rotation_2020} (F20).
  \label{fig:litcomp}
	}
\end{figure*}

From their rotation period analysis of \cn,
\citet{fritzewski_rotation_2020} demonstrated that between 100 and
200\,Myr, young clusters (\cn, Blanco\,1, Psc-Eri, M\,35, M\,50, and
the Pleiades) all show similar rotation period distributions.  The FGK
stars show a ``slow sequence'', a population of ``rapid rotators'',
and in some cases stars in the ``void'' between the two populations
({\it e.g.,}, Psc-Eri and \cn\ show more stars in the void than the
Pleiades).  The late M dwarfs with detected rotation periods tend to
rotate rapidly ($<3$\,days), though there may also be late M dwarfs
with slow ($>20$\,days) rotation periods
\citep{stauffer_rotation_2016}.  Although these late M dwarfs are
beyond the magnitude limit of our analysis, they have been explored in
depth by \citet{Irwin_NGC2516_2007} and
\citet{fritzewski_rotation_2020}.  A comparison of our results is
shown in Figure~\ref{fig:litcomp}.  The bottom panel of this plot
includes, in order of precedence: rotation periods from our analysis,
from ``Class 1 and 2'' \citet{fritzewski_rotation_2020} rotators, and
from the rotators in \citet{Irwin_NGC2516_2007}.  The main
contribution of our study beyond these previous analyses is our
sensitivity across both the core and halo of \cn, and a corresponding
expansion in the number of stars that can be analyzed.  In the
following section, we therefore focus on the origins of the features
in the rotation-color diagram, and whether our expanded sample can
provide any new insights.

@@

The other features of the rotation-color diagram are indications that
magnetic braking is not always the only important effect.  The
question of why the rapid rotators exist, for instance, requires
additional physics.  Models that vary core-envelope decoupling
timescales and pre-MS disk lifetimes can reproduce each type of
population
\citep{Irwin_NGC2516_2007,gallet_improved_2013,gallet_improved_2015};
the question then shifts to why these timescales should differ for
stars that are otherwise identical.  Another possibility
is that proposed by \citet{matt_mass-dependence_2015}: the
saturation of the magnetic dynamo yields ${\rm d}P/{\rm d}t \sim P$ when $P >
P_{\rm sat}$, ${\rm d}P/{\rm d}t \sim P^{-1}$ for $P < P_{\rm sat}$, and a
maximal spin-down rate in the intermediate regime.  This could
naturally produce the period bimodality observed at ages of
$\approx$100\,Myr (though it fails to explain the widening of the
distribution observed at later times
\citealt{godoyrivera_stellar_2021}).  Other significant processes
could be external to the star entirely: \citet{qureshi_signature_2018}
for instance have reported that mergers between giant planets and the
star could explain a non-neglible fraction of the rapid rotators.

@@

On the other hand, our ability to reliably identify members of these
halos is rather surprising.  These stars have remained hidden in the
background of the Galaxy throughout centuries of modern astronomy.
Their discovery represents an important step toward understanding
the dispersal of open clusters.
Combined with wide-field photometric surveys, the stars themselves
enable advances in studies of stellar
angular momentum evolution.
Combined with wide-field spectroscopic
surveys \citep[{\it e.g.},][]{kollmeier_2017}, the halos could also
improve our
understanding of the chemical evolution of low-mass stars.
Finally, the halo stars provide an expanded sample of
age-dated stars to search for planets \citep[{\it
e.g.},][]{newton_2021}, which may help in benchmarking our understanding of
planetary evolution.

@@

The authors are indebted, in reverse-alphabetical order, to G.~Zhou,
B.~Tofflemire, A.~McWilliam, E.~Newton, L.~Hillenbrand, A.~Kraus, and
K.~Hawkins for the discussions on young stars, rotation, and lithium
that encouraged this analysis.
%
L.G.B. and J.H. acknowledge support by the TESS GI Program, programs
G011103 and G022117, through NASA grants 80NSSC19K0386 and
80NSSC19K1728.
%
L.G.B. was also supported by a Charlotte Elizabeth Procter Fellowship
from Princeton University.
%
This study was based in part on observations at Cerro Tololo
Inter-American Observatory at NSF's NOIRLab (NOIRLab Prop{.} ID
2020A-0146; 2020B-0029 PI: L{.}~Bouma), which is managed by the
Association of Universities for Research in Astronomy (AURA) under a
cooperative agreement with the National Science Foundation.
%
% ACKNOWLEDGE PFS / CAMPANAS.
%
This paper also includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
We thank the TESS Architects (G.~Ricker, R.~Vanderspek, D.~Latham,
S.~Seager, J.~Jenkins) and the many TESS team members for their
efforts to make the mission a continued success.
%
